Constructive forcing

### 2023-09-15 slides-herrsching2023.tex
More notes

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\centering\includegraphics[height=\paperheight]{multiverse-faded-169}\end{minipage}}
\begin{frame}[c]
  \jnote{1}{These are the slides for a three-part lecture series for the autumn school
  \fixedhref{https://www.mathematik.uni-muenchen.de/~schwicht/pc23.php}{\emph{Proof and Computation}}
  organized by Klaus Mainzer, Peter Schuster and Helmut Schwichtenberg, held in
  Herrsching from September 20th to September 16th, 2023.}

@@

\newcommand\ytl[2]{%
  \parbox[b]{4.5em}{\hfill{#1}~$\cdots\cdots$~}%
  \makebox[0pt][c]{\color{mypurple}$\bullet$}{\color{mypurple}\vrule}\quad%
  \parbox[c]{9.3cm}{\vspace{7pt}\raggedright#2\\[7pt]}%
  \\[-2.0pt]}
\begin{frame}{A brief timeline}
  \jnote{4}{
    Gödel's proof is by the \hil{$L$-translation}, where $L$ is the
    ``constructible universe''. This translation ``relativizes quantification
    to~$L$'', for instance the~$L$-translation of
    \[ \varphi \defeqv \bigl(\forall x\_ \exists y\_ \ldots\bigr) \qquad\text{is}\qquad
    \varphi^L \equiv \bigl(\forall(x \in L)\_ \exists(y \in L)\_ (\ldots)^L\bigr). \]
    We can then verify, in a weak metatheory such as~\textsc{pra}, that for
    every formula~$\varphi$ in the language of set theory,
    \[ \text{if $\textsc{zfc} \vdash \varphi$, then $\textsc{zf} \vdash
    \varphi^L$}. \]
    Specializing to~$\varphi \defeqv \bot$ we obtain: If~\textsc{zfc} is
    inconsistent, then so is~\textsc{zf}. The axiom of choice does not
    introduce new inconsistencies.

    In modern semantic language: While the axiom of choice might fail in the
    base universe~$V$ (= the class of all sets), it always holds in~$L$.
    Gödel's~$L$ was the first \emph{inner model} (= class-sized model of set
    theory) explicitly studied, nowadays we know many.
  }

@@

    Just as a given ring~$R$ or group can be extended in various ways, to include
    ``generic elements'' as in~$R[X]$ or elements with prescribed relations as
    in~$R[X]/(X^2+1) =\vcentcolon R[i]$. The idea of forcing is to construct
    similar such extensions, but of universes (traditionally set-sized models
    of~\textsc{zfc}, but also class-sized models of~\textsc{zfc}, or models
    of intuitionistic set theories, or models of type theories, or even models
    of arithmetic) instead of rings.

    In semantic language, from a high level the idea is the following: Whether
    the base universe~$V$ contains a cardinal number intermediate
    between~$\aleph_0$ and~$2^{\aleph_0}$ is uncertain. But there is a certain
    extension of the base universe---constructed by forcing---which does
    contain such a number. Like the base~$V$, this forcing extension still
    validates the axioms of~\textsc{zfc}. Hence there cannot be a
    \textsc{zfc}-proof of~\textsc{ch}, as in Cohen's extension~$\neg\textsc{ch}$
    holds.

    Syntactically, Cohen's forcing provides us with an explicit formula
    translation~$\varphi \mapsto \varphi^C$ such that~\textsc{pra} proves:
    For every formula~$\varphi$, if~$\textsc{zfc}{\text{+}}\neg\textsc{ch} \vdash \varphi$
    then~$\textsc{zfc} \vdash \varphi^C$.
    \bigskip
  }

@@

\begin{frame}{What can forcing do for you?}
  \vspace*{-0.5em}

>>>>

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\centering\includegraphics[height=\paperheight]{multiverse-faded}\end{minipage}}
\begin{frame}[c]
  \jnote{1}{
    These are the slides for a three-part lecture series for the autumn school
    \fixedhref{https://www.mathematik.uni-muenchen.de/~schwicht/pc23.php}{\emph{Proof and Computation}}
    organized by Klaus Mainzer, Peter Schuster and Helmut Schwichtenberg, held in
    Herrsching from September 20th to September 16th, 2023.

    We explore, in a constructive metatheory, the ``Kripke--Joyal semantics of
    the internal language of Grothendieck toposes over sites given by
    preorders'' together with applications in constructive algebra and
    combinatorics, but without presupposing familiarity with or using language
    from topos or category theory.
  }

@@

\newcommand\ytl[2]{%
  \parbox[b]{4.5em}{\hfill{#1}~$\cdots\cdots$~}%
  \makebox[0pt][c]{\color{mypurple}$\bullet$}{\color{mypurple}\vrule}\quad%
  \parbox[c]{9.3cm}{\vspace{7pt}\raggedright#2\\[7pt]}%
  \\[-2.0pt]}
\begin{frame}{A brief timeline}
  \jnote{4}{
    Gödel's proof is by the \hil{$L$-translation}, where $L$ is the
    ``constructible universe''. This translation ``relativizes quantification
    to~$L$'', for instance the~$L$-translation of
    \[ \varphi \defeqv \bigl(\forall x\_ \exists y\_ \ldots\bigr) \qquad\text{is}\qquad
    \varphi^L \equiv \bigl(\forall(x \in L)\_ \exists(y \in L)\_ (\ldots)^L\bigr). \]
    We can then verify, in a weak metatheory such as~\textsc{pra}, that for
    every formula~$\varphi$ in the language of set theory:
    If $\textsc{zfc}\text{+}\textsc{ch} \vdash \varphi$, then $\textsc{zf} \vdash
    \varphi^L$.

    Specializing to~$\varphi \defeqv \bot$ we obtain in particular: If~\textsc{zfc} is
    inconsistent, then so is~\textsc{zf}. The axiom of choice does not
    introduce new inconsistencies.

    In modern semantic language: While the axiom of choice and~\textsc{ch} might fail in the
    base universe~$V$ (= the class of all sets), they always hold in~$L$.
    Gödel's~$L$ was the first \emph{inner model} (= class-sized model of set
    theory) explicitly studied, nowadays we know many.
  }

@@

    Recall that a given ring~$R$ or group can be extended in various ways, to include
    ``generic elements'' as in~$R[X]$ or elements with prescribed relations as
    in~$R[X]/(X^2+1) =\vcentcolon R[i]$. The idea of forcing is to construct
    similar such extensions, but not of rings but of universes (traditionally set-sized models
    of~\textsc{zf} or~\textsc{zfc}, but also class-sized models of~\textsc{zfc}, or models
    of intuitionistic set theories, or models of type theories, or even models
    of arithmetic).

    In semantic language, from a high level the idea of Cohen's independency proof is the following: Whether
    the base universe~$V$ contains a cardinal number intermediate
    between~$\aleph_0$ and~$2^{\aleph_0}$ is uncertain. But there is a certain
    extension of the base universe---constructed by forcing---which does
    contain such a number. Like the base~$V$, this forcing extension still
    validates the axioms of~\textsc{zfc}. Hence there cannot be a
    \textsc{zfc}-proof of~\textsc{ch}, as in Cohen's extension~$\neg\textsc{ch}$
    holds.

    Syntactically, Cohen's forcing provides us with an explicit formula
    translation~$\varphi \mapsto \varphi^C$ such that~\textsc{pra} proves:
    For every formula~$\varphi$, if~$\textsc{zfc}{\text{+}}\neg\textsc{ch} \vdash \varphi$,
    then~$\textsc{zfc} \vdash \varphi^C$.
    \bigskip
  }

  \jnote{6}{
    Joel David Hamkins argues: In view of our rich experience with worlds which
    validate~\textsc{ch} and worlds which don't, we shouldn't be surprised that
    no proposed new axiom for settling~\textsc{ch} is ultimately convincing.

    Instead, we should embrace the multiverse of all models of set theory and
    explore how the truth value of statements of interest change when we
    travel the multiverse (for instance, by passing from some universe to one
    of its forcing extensions).

    In this generalized sense, the continuum hypothesis is settled: We have a
    good understanding of the stability properties of~\textsc{ch} under
    important constructions. In particular, for a certain precise meaning of
    ``universe'' and ``extension'', we know that~\textsc{ch} is a
    \emph{switch}: $\necessary(\possible\textsc{ch} \wedge
    \possible\neg\textsc{ch})$; in words: Every universe can be extended both to a
    universe in which~\textsc{ch} holds and to a universe in
    which~$\neg\textsc{ch}$ holds.

    An exposition and references for further reading about the multiverse
    position can be found
    \fixedhref{https://www.speicherleck.de/iblech/stuff/multiverse.pdf}{here}.
  }

  \jnote{8}{
    J. Roitman, The uses of set theory, \emph{Math.\@ Intelligencer}
    \textbf{14}(1) (1992), 63--69.

    Forcing is useful not only to explore the range of foundational
    possibility; it has many more applications across several subjects of
    mathematics.

    In particular, we will discuss applications of the constructive version of
    classical set-theoretic forcing in constructive algebra and combinatorics.
  }

@@

\begin{frame}{What can forcing do for you?}
  \jnote{1}{
    Forcing~$\neg\textsc{ch}$ was the historical use case for forcing.

    With constructive forcing, where we do not blanketly assume that the base
    universe validates~\textsc{lem}, we can inquire the status of~\textsc{lem};
    a result in this direction is that~\textsc{lem} is, like~\textsc{ch} in
    the context of~\textsc{zfc}, a \emph{switch}:

    The base universe can always be extended in such a way as to
    force~$\neg\textsc{lem}$ (in fact, most forcing extensions will
    falsify~\textsc{lem} even in case the base universe validates it), and also
    in such a way to force~\textsc{lem}. The latter provides us with a semantic
    view of one of the techniques for extracting constructive content from
    classical proofs, namely the
    \fixedhref{https://www.speicherleck.de/iblech/stuff/slides-fischbachau2022.pdf}{double-negation
    translation combined by the continuation trick}.
  }

  \jnote{2}{
    We will discuss the particular example of the (naive formulation of the)
    fundamental theorem of algebra below.

    Forcing has been used to construct countermodels to various questions of
    constructive (reverse) mathematics, too many to list here. To give just one
    pointer, a countermodel for the classical implication ``if there is no
    infinite descending chain, then the partial order is inductively
    well-founded'' is presented here:
    A. Blass, Well-ordering and induction in intuitionistic logic and topoi,
    in: \emph{Mathematical Logic and Theoretical Computer Science}. Ed.\@ by
    D.\@ Kueker, E.\@ Lopez-Escobar, and C.\@ Smith. Vol.\@ 106. Lect.\@ Notes
    Pure Appl. Math. Marcel Dekker, 1987, pp.\@ 29--48.
  }

  \jnote{3}{
    The real numbers don't contain a number~$i$ such that~$i^2 = 0$. Still, for
    many results in real analysis, it is convenient to broaden our notion of
    existence and pass to the complex numbers; the imaginary unit is a
    \emph{mathematical phantom} in the
    \fixedhref{http://www.wra1th.plus.com/gcw/math/MathPhant.html}{sense of
    Gavin Wraith}, a useful tool helping us deduce results about real numbers.
    Nowadays there are few ontological concerns about the imaginary unit: We
    understand that, in the end, every statement about complex numbers can be
    recast as a statement about pairs of real numbers.

    In exactly the same fashion, the objects furnished by forcing can be
    understood as useful fictions. We will discuss how statements about
    the forcing extension can be recast as statements about the base universe.

    Given an inhabited set~$X$ in the base universe, most pronouncedly a set
    which is uncountable or for which no surjection~$\NN \twoheadrightarrow X$
    can be efficiently evaluated, a particularly tantalizing fiction is the
    \emph{generic surjection}~$\NN \twoheadrightarrow X$. It exists in a
    custom-tailored forcing extension of the base universe and is useful to
    apply tools made for the countable setting to the uncountable; we will
    discuss an example on the next slide.
  }

  \jnote{4}{
    Typically there are already plenty of maps~$\NN \to X$ in the base
    universe; hence constructing a forcing extension which contains a ``fresh''
    such map---the so-called \emph{generic sequence}---is not something which
    would usually be contemplated in classical set-theoretic forcing.

    In the context of constructive mathematics, however, the generic sequence
    turns out to be quite useful. It can be used to cast in a familiar naive
    language---the language of infinite sequences---definitions, results and
    proofs which use constructively more appropriate inductively defined
    notions. We will discuss an example below.
  }

  \jnote{6}{
    Slides by Matthias Hutzler:
    \fixedhref{https://matthias-hutzler.de/pc23/slides.pdf}{Introduction to
    synthetic algebraic geometry}
  }

  \vspace*{-0.5em}
